---
title: "Human GLORI site annotation and matched background construction"
author: "posMatchR"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Human GLORI site annotation and matched background construction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 4
)
```

# Overview

This vignette demonstrates a minimal `posMatchR` workflow using a 25,000 m6A site human GLORI HEK293T `GRanges` object. The analysis annotates the sites with transcript context, extracts centred 5-mers, builds a candidate background universe from the GLORI-observed 5-mers with at least 100 foreground instances, and performs exact centred 5-mer matching.

# Resources

```{r packages}
library(posMatchR)

txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene::TxDb.Hsapiens.UCSC.hg38.knownGene
genome <- BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38
orgdb <- org.Hs.eg.db::org.Hs.eg.db
human_chrs <- paste0("chr", c(1:22, "X", "Y"))
```

This vignette uses a precomputed transcript-resource object for the human hg38 knownGene TxDb. This keeps package checks fast; the equivalent object can be rebuilt with `build_tx_resources(txdb)` or with the script in `inst/scripts/make_human_glori_tx_resources_cache.R`.

```{r resources}
resource_file <- system.file(
    "extdata", "hg38_knownGene_tx_resources_GLORI_demo.rds",
    package = "posMatchR",
    mustWork = TRUE
)
resources <- readRDS(resource_file)
```

# Load and annotate GLORI sites

The example data file contains one `GRanges` object named `GLORI`. `load_sites()` loads the object and prepares it as width-one sites.

```{r load-sites}
glori_file <- system.file(
    "extdata", "GLORI_HEK293T_25K.Rdat",
    package = "posMatchR",
    mustWork = TRUE
)

sites <- load_sites(glori_file)
sites <- sites[seq_len(min(length(sites), 5000L))]
sites
```

Adds annotations to the single-width GRange object.

```{r annotate-sites}
sites <- annotate_sites(
    gr = sites,
    txdb = txdb,
    resources = resources,
    seqstyle = "UCSC",
    chrs = human_chrs,
    tx_select = "longest",
    orgdb = orgdb,
    gene_keytype = "ENTREZID",
    gene_symbol_col = "SYMBOL",
    gene_name_col = "GENENAME",
    drop_unannotated = TRUE
)

sites <- add_kmer(
    gr = sites,
    genome = genome,
    k = 5,
    out_col = "kmer",
    seqstyle = "UCSC",
    chrs = human_chrs
)

head(as_basic_site_table(sites))
```

# Annotation plots

```{r annotation-plots}
plot_metagene_density(sites)
plot_junction_distance_density(sites)
plot_kmer_counts(sites, top_n = 15)
```


Stop here if only interested in annotations and a background set is not required.

# Build and match a 5-mer background

Candidate background sites are generated from the GLORI-observed centred 5-mers that occur at least 100 times in the foreground. The matched background requires exact centred 5-mer agreement between each retained positive site and its selected negative site.

```{r universe}
background <- make_kmer_universe(
    foreground = sites,
    txdb = txdb,
    genome = genome,
    kmer_col = "kmer",
    min_count = 100,
    scope = "transcripts",
    regions = c("fiveUTR", "coding", "threeUTR"),
    resources = resources,
    exclude_foreground = TRUE
)

# annotate all sites
combined <- combine_site_sets(sites, background)
combined <- annotate_sites(
    gr = combined,
    txdb = txdb,
    resources = resources,
    seqstyle = "UCSC",
    chrs = human_chrs,
    tx_select = "longest",
    orgdb = orgdb,
    gene_keytype = "ENTREZID",
    gene_symbol_col = "SYMBOL",
    gene_name_col = "GENENAME",
    drop_unannotated = TRUE
)
combined <- add_kmer(combined, genome = genome, k = 5, seqstyle = "UCSC", chrs = human_chrs)
```

The following generates a background set based on matched metagene and geometry features.

```{r build-background}
matched <- match_background(
    gr = combined,
    kmer_match = TRUE,
    kmer_col = "kmer",
    bin_match = TRUE
)
matched_pairs <- subset_matched_sets(matched)
matched_pairs
```

# Matched-set plots

```{r matched-plots}
plot_metagene_density(matched_pairs)
plot_junction_distance_density(matched_pairs)
plot_kmer_counts(matched_pairs, top_n = 15)
head(summarise_matched_kmer_balance(matched_pairs))
```

```{r matching-diagnostic}
plot_matching_pca(matched, max_unselected = 5000)
```

# Session information

```{r session-info}
sessionInfo()
```
